Missing data: Current practice in football research and recommendations for reporting missingness

### 2020-09-03 paper/paper.Rmd
add changes from meeting

Missing data are often unavoidable. The reason values go missing, along
with decisions made of how missing data are handled (deleted or
imputed), can have a profound effect on the validity, usability and
accuracy of a study's results. In this article, we aimed to: estimate
the proportion of studies in football research that included a missing
data statement, highlight several improvable practices in relation to
missing data, and provide recommendations for exploring, visualising and
reporting missingness. Football related articles, published in 2019 were
studied. A survey of 137 articles, sampled at random, was conducted to
determine whether a missing data statement was included. As expected, the
proportion of studies in football research that included a missing data
statement was low, at only X% (95% confidence interval; X% to X%); suggesting that 
missingness is seldom considered by researchers. We suspect that this result 
is consistent with the wider sports science literature. We recommend 
researchers describe the number and percentage of missing
values, including when there are no missing values. Exploratory analysis
should be conducted to explore missing values, and visualisations
describing missingness overall should be provided in the paper, or at
least supplementary materials. Missing values should almost always be
imputed, and imputation methods should be explored to ensure they are
appropriately representative.

**Keywords:** Exercise, imputation, missingness, missings, naniar, sport

@@

Missing data are values that should have been observed, but were not. Since the
values are unobserved, this can mean the population of interest might not be 
properly sampled, inducing bias into study results. Left undetected or poorly handled, missing data can undermine the validity and accuracy of research results
[@Sainani2015; @Nakagawa2008; @Sterne2009]. Data can go
missing many ways. For example, accidentally skipping a survey question,
equipment failure, or intentionally not recording values. Before
analysis, missing data must be handled, with values typically deleted or
imputed. There is no universal approach to handling missing data. 
Contextual factors--such as, the study design and objective, and
pattern of missingness--determine how missings should be handled, on a
case-by-case basis [@Sainani2015].
The effect of missing data on the validity and accuracy of a study's
results is largely determined by the reasons for missingness, and how
missing data are handled.

A missing value should have been observed or measured in the data, but was not. For example, a goal scored in football takes place, but the scorekeeper wasn't paying attention and did not record an event. This "goal" is now a missing value. Compare this to a game being rained out and no game taking place. There is no missing data because there is no data to be observed. A final example - a study screens out individuals who do not meet criteria for the study. These are not missing values, because there was no intent for these to be observed or measured.

@@

It is imperative that missing data, or lack thereof, are reported [@Schafer2002]. 
While there has been significant interest in modelling missing
data [@refs], the exploration and reporting of missings have received
less attention [@Schafer2002]. This article aimed to: (1) estimate the proportion
of articles that report missing data in football research; (2) highlight
several improvable missing data practices; and (3) provide recommendations for
exploring and reporting missingness.

@@

To estimate the proportion of articles that report missing data in
football research, we conducted a systematic search [@PRISMA]. Football related
articles published in 2019 were studied. Three major sports medicine
databases (SPORTDiscus, Embase and Cinahl) were searched, using key
terms and search limits (Supplement 2). Included articles were: (a) written in English; (b) had an accessible
full-text; and (c) included quantitative data. The search process is
summarised in the PRISMA figure provided in Supplement 2.

Our intention was to survey, at random, 10% of the total articles found
(after title and abstract screening). We were primarily interested in
whether missing data were reported or acknowledged; and if not, whether
a dataset was shared. Where applicable, information about how missing
values were handled was also extracted (see Supplement 2). Two authors
independently extracted the data (X% agreement). Findings are reported
as the proportion and 95% confidence interval. Supplement 2 provides
further details of the search.

>>>>

Missing data are often unavoidable. The reason values go missing, along
with decisions made of how missing data are handled (deleted or
imputed), can have a profound effect on the validity, usability and
accuracy of a study's results. In this article, we aimed to: estimate
the proportion of studies in football research that included a missing
data statement, highlight several improvable practices in relation to
missing data, and provide recommendations for exploring, visualising and
reporting missingness. Football related articles, published in 2019 were
studied. A survey of 137 articles, sampled at random, was conducted to
determine whether a missing data statement was included. As expected, the
proportion of studies in football research that included a missing data
statement was low, at only X% [95% confidence interval; X% to X%]; suggesting that 
missingness is seldom considered by researchers. We suspect that this result 
is consistent with the wider sports science literature. We recommend 
researchers describe the number and percentage of missing
values, including when there are no missing values. Exploratory analysis
should be conducted to explore missing values, and visualisations
describing missingness overall should be provided in the paper, or at
least supplementary materials. Missing values should almost always be
imputed, and imputation methods should be explored to ensure they are
appropriately representative.

**Keywords:** Imputation, missingness, missings, naniar, sport, football

@@

Missing data are values that should have been observed, but were not. Since the
values are unobserved, this can mean the population of interest might not be 
properly sampled, inducing bias into study results. Left undetected or poorly handled, missing data can undermine the validity and accuracy of research results
[@Sainani2015; @Nakagawa2008; @Sterne2009]. Data can go
missing many ways. For example, accidentally skipping a survey question,
equipment failure, or intentionally not recording values. Before
analysis, missing data must be handled, with values typically deleted or
imputed. There is no universal approach to handling missing data. 
Contextual factors--such as, the study design and objective, and
pattern of missingness--determine how to handle missings on a
case-by-case basis [@Sainani2015].

For clarity, we describe some examples of missingness. An Australian football data collector was taking lunch during the third quarter, so no data was recorded for this time period. Compare this to an AFL game being cancelled due to COVID19. There is no missing data because there is no data to be observed. If a study screens out individuals who do not meet inclusion criteria, this is not a missing value, because there is no intent for these values to be observed or measured.

@@

It is imperative that missing data (or lack thereof --Absence of
evidence is not evidence of absence) are reported. However, outside of
sports science, researchers seldom consider missing data [@Schafer2002] (<< NOTE - I'm not sure we mean that?).
Moreover, while there has been significant interest in modelling missing
data [@refs], the exploration and reporting of missing data have received
less attention [@Schafer2002]. This article aimed to: (1) estimate the proportion
of articles that report missing data in football research; (2) highlight
several bad missing data practices; and (3) provide recommendations for
exploring and reporting missingness.

@@

To estimate the proportion of articles that report missing data in
football research, we conducted a systematic search [@Schafer2002]. Football related
articles published in 2019 were studied. Three major sports medicine
databases (SPORTDiscus, Embase and Cinahl) were searched, using key
terms and search limits (Supplement 1). Included articles were: (a)
published in 2019; (b) written in English; (c) had an accessible
full-text; and (d) included quantitative data. The search process is
summarised in Figure 1.

```{r prisma-example, eval = FALSE}
require(PRISMAstatement)

prisma()

[[https://cran.r-project.org/web/packages/PRISMAstatement/vignettes/PRISMA.html]{.ul}](https://cran.r-project.org/web/packages/PRISMAstatement/vignettes/PRISMA.html)

library(PRISMAstatement)

e.g.,

prisma(found = x,

found_other = x,

no_dupes = x,

screened = x,

screen_exclusions = x,

full_text = x,

full_text_exclusions = x,

qualitative = x,

quantitative = x,

width = 800, height = 800)
```

+---------------------------------+-----------------------------------+
| Imported                        | 3,863                             |
+=================================+===================================+
| After duplicates removed        | 1,954                             |
+---------------------------------+-----------------------------------+
| After title and abstract screen | 1,373                             |
+---------------------------------+-----------------------------------+
| Random sample full-text         | 199 (needed for 10% sample)       |
+---------------------------------+-----------------------------------+
| Full-text included              | 137                               |
+---------------------------------+-----------------------------------+
| Full-text excluded              | Full-text papers excluded (n=62)  |
|                                 | with reasons:                     |
|                                 |                                   |
|                                 | Qualitative study (n=)            |
|                                 |                                   |
|                                 | Not in english (n=)               |
|                                 |                                   |
|                                 | Unable to access full-text (n=)   |
+---------------------------------+-----------------------------------+

**Figure 1.** PRISMA flow chart.

Our intention was to survey, at random, 10% of the total articles found
(after title and abstract screening). We were primarily interested in
whether missing data were reported or acknowledged; and if not, whether
a dataset was shared. Where applicable, information about how missing
values were handled was also extracted (see Supplement 1). Two authors
independently extracted the data (X% agreement). Findings are reported
as the proportion and 95% confidence interval. Supplement 1 provides
further details of the search.
